---
title: "02_visualise"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# Rscript -e "rmarkdown::render('src/02_visualise.Rmd', output_dir = 'out/analysis/')" # nolint: line_length_linter.

# knitr opts
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)

# libraries
library(magrittr)
library(ggplot2)
source("../bin/values_to_colors.R")
```

# Load SNVs and indels

We load the matrix of variant counts from the `Sequoia` output. This includes 
all indels and SNVs after filtering germline and artifactual variants.

```{r load_muts}
nv_filtered <-
  read.table("out/PD56034/sequoia/PD56034_both_NV_filtered_all.txt") %>%
  tibble::as_tibble(rownames = "var") %>%
  tidyr::separate_wider_delim(cols = "var", delim = "_",
                              names = c("chr", "pos", "ref", "alt")) %>%
  # classify variant types
  dplyr::mutate(
    var_type = dplyr::case_when(nchar(ref) + nchar(alt) > 2 ~ "indel_count",
                                nchar(ref) + nchar(alt) == 2 ~ "snv_count")) %>%
  # binarise counts
  dplyr::mutate(dplyr::across(where(is.numeric), ~ as.numeric(.x > 0))) %>%
  # get burdens
  dplyr::group_by(var_type) %>%
  dplyr::summarise(dplyr::across(where(is.numeric), sum)) %>%
  # get one row per colony
  tidyr::pivot_longer(cols = -var_type, names_to = "colony_id",
                      values_to = "count") %>%
  tidyr::pivot_wider(names_from = var_type, values_from = count) %>%
  # remove normal
  dplyr::filter(colony_id != "normal")
```

# Load colony metadata and CanApps QC

```{r load_metadata}
# load QC from canapps
colony_seq_qc <-
  readxl::read_xls(
    path = "data/Cancer_Pipeline_Reports_SampleDetails.xls",
    skip = 2
  ) %>%
  dplyr::transmute(colony_id = Sample, sample_coverage = `Seq X`)

# load colony metadata, add QC
colony_metadata <-
  readr::read_tsv("out/metadata/colony_metadata.tsv") %>%
  # add coverage
  dplyr::inner_join(colony_seq_qc) %>%
  # get well row and column ids
  tidyr::separate_wider_position(
    cols = "well_id",
    widths = c("well_column_id" = 1, "well_row_id" = 2),
    cols_remove = FALSE
  ) %>%
  # add burdens
  dplyr::left_join(nv_filtered)
```

# Plot burden and coverage

## Coverage by plate

```{r plot_cov_by_plate_bar}
# colour by timepoint x plate id, ordered by row
colony_metadata %>%
  dplyr::mutate(colony_id = forcats::fct_reorder(colony_id, well_row_id)) %>%
  ggplot(aes(x = colony_id, y = sample_coverage, fill = colony_timepoint)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ggh4x::facet_nested(~ colony_timepoint + plate_id,
    scales = "free_x",
    space = "free_x"
  ) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_blank())
```

While the two post-chemotherapy plates show fairly consistent coverage across 
colonies, the pre-chemotherapy plate has a few colonies with very high coverage,
while most colonies have ~ half the coverage of post-chemotherapy.

## Burden by plate

```{r plot_burden_by_plate_bar}
# colour by timepoint x plate id, ordered by row
colony_metadata %>%
  dplyr::mutate(colony_id = forcats::fct_reorder(colony_id, well_row_id)) %>%
  tidyr::pivot_longer(cols = c("snv_count", "indel_count"),
                      names_to = "var_type", values_to = "count") %>%
  ggplot(aes(x = colony_id, y = count, fill = colony_timepoint)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ggh4x::facet_nested(var_type ~ colony_timepoint + plate_id,
    scales = "free",
    space = "free_x"
  ) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_blank())
```

We can see that this coverage pattern is also reflected in the resultant 
burdens, both for SNVs and indels.

## Coverage by plate well

```{r plot_cov_by_well_tile}
colony_metadata %>%
  ggplot(aes(x = well_column_id, y = well_row_id, fill = sample_coverage)) +
  geom_tile() +
  ggh4x::facet_nested(~ plate_id + colony_timepoint) +
  theme_classic()
```

Inspecting the coverage plate-x-well indicates that those 10 colonies with very
high coverage in the pre-chemotherapy plate are all located consecutively, in 
wells A01 to B02. This suggests that the strong change in coverage is the 
result of an experimental error.

## Burden by plate well

```{r plot_burden_by_well_tile}
# snvs
colony_metadata %>%
  ggplot(aes(x = well_column_id, y = well_row_id, fill = snv_count)) +
  geom_tile() +
  ggh4x::facet_nested(~ plate_id + colony_timepoint) +
  theme_classic()
# indels
colony_metadata %>%
  ggplot(aes(x = well_column_id, y = well_row_id, fill = indel_count)) +
  geom_tile() +
  ggh4x::facet_nested(~ plate_id + colony_timepoint) +
  theme_classic()
```

This is again reflected in the SNVs and indel burdens. This suggests that the 
burdens are strongly influenced by this technical variate, rather than a true 
biological difference.

# Plot burdens vs coverage

To be sure, we review the relationship between coverage and burden in SNVs and 
indels.

```{r plot_burdens_vs_cov, fid.width = 10}
colony_metadata %>%
  tidyr::pivot_longer(cols = c("snv_count", "indel_count"),
                      names_to = "var_type", values_to = "count") %>%
  ggplot(aes(x = sample_coverage, y = count, color = colony_timepoint)) +
  geom_point() +
  facet_wrap(~ var_type, scales = "free_y") +
  theme_classic()
```

There appears to be a strong linear relationship between coverage and mutation 
burden in both SNVs and indels. Additionally, most of the post-chemotherapy 
colonies have a higher sample coverage than the pre-chemotherapy colonies, with
a few outlying pre-chemotherapy colonies with very high coverage, and 
proportionately higher mutational burdens. This reflects what was previously
observed.

# Plot trees

The topology and branch length of these trees makes sense in light of these 
findings. The mutation burdens in the pre-chemotherapy colonies are much lower 
than would be expected given the donor age, except for a few of the tips. This 
is due to the low coverage.

```{r plot_trees, fig.height = 10}
colony_metadata %>%
  {
    split(., .$donor_id)
  } %>%
  purrr::map2(., names(.), function(donor_metadata, donor_id) {
    files <-
      file.path("out", donor_id, "sequoia") %>%
      list.files(pattern = "\\.tree$", full.names = TRUE)
    if (length(files) > 0) {
      files %>%
        purrr::map(function(file) {
          # load tree
          tree_obj <- ape::read.tree(file)

          # get rownames
          dm <- tibble::column_to_rownames(donor_metadata, "colony_id")

          # get metadata variables with multiple values
          colours <-
            dm[tree_obj$tip.label, ] %>%
            dplyr::select(-well_id) %>%
            purrr::map(~ length(unique(na.omit(.x)))) %>%
            {
              .[. > 1]
            }  %>%
            purrr::map2(., names(.), function(colour_var_n, colour_var_name) {
              # get palette
              colours <- dm[tree_obj$tip.label, colour_var_name]
              colours <- colours[!is.na(colours)]
              if (is.numeric(colours)) {
                pal <-
                  values_to_colors(colours,
                    min_val = min(colours),
                    max_val = max(colours)
                  )
                pal <- pal %>% purrr::set_names(colours)
              } else {
                if (colour_var_n == 2) {
                  pal <- c("red", "blue")
                } else {
                  pal <- c(RColorBrewer::brewer.pal(colour_var_n, "Set3"))
                }
                pal <- pal %>% purrr::set_names(unique(colours))
              }
              tree_colours <-
                pal[dm[tree_obj$tip.label, colour_var_name]]

              # plot
              plot(tree_obj,
                tip.color = tree_colours, cex = 0.5,
                main = basename(file)
              )
              ape::axisPhylo(side = 1, backward = FALSE)
              legend("topright",
                legend = unique(names(tree_colours)),
                col = unique(tree_colours), lty = 1:2, cex = 0.8,
                title = colour_var_name
              )
            })
        })
    }
  })
```